---
title: "Microarray Analysis Pipeline"
subtitle: "Reproducible miRNA and Clariom S/mRNA workflows"
format:
  html:
    toc: false
    number-sections: true
    smooth-scroll: true
    code-fold: false
    embed-resources: true
execute:
  echo: true
  warning: true
  message: true
---

<style>
body {
  line-height: 1.6;
}
h1 {
  border-bottom: 3px solid #2f5d50;
  padding-bottom: 0.45rem;
  margin-top: 2.8rem;
}
h2 {
  border-left: 6px solid #6aa88f;
  padding-left: 0.75rem;
  margin-top: 2.2rem;
}
h3 {
  color: #2f5d50;
  margin-top: 1.6rem;
}
table {
  border: 1px solid #c8d8d1;
  border-radius: 6px;
  overflow: hidden;
}
thead th {
  background: #eaf4ef;
  color: #21493c;
}
blockquote {
  background: #f1f8f5;
  border-left: 5px solid #4c9a78;
  padding: 0.8rem 1rem;
  color: #23463b;
}
pre.sourceCode {
  border-left: 4px solid #7aa9c2;
  border-radius: 5px;
}
.workflow-box,
.purpose-box,
.expected-box {
  border: 1px solid #b7d3c7;
  border-radius: 8px;
  padding: 1rem 1.25rem;
  margin: 1.25rem 0;
  background: #f6fbf8;
}
.workflow-box {
  border-left: 7px solid #4c9a78;
  background: #eef8f2;
}
.purpose-box {
  border-left: 7px solid #7aa9c2;
  background: #f1f7fa;
}
.expected-box {
  border-left: 7px solid #d39a3a;
  background: #fff9ed;
}
hr.section-divider {
  border: 0;
  border-top: 4px solid #333;
  margin: 3rem 0 2.5rem;
}
.downstream-file code {
  color: #145c32;
  background-color: #e7f6ec;
  border: 1px solid #86c89d;
  border-radius: 4px;
  padding: 0.15rem 0.35rem;
  font-weight: 700;
}
.downstream-badge {
  display: inline-block;
  color: #145c32;
  background-color: #e7f6ec;
  border: 1px solid #86c89d;
  border-radius: 999px;
  padding: 0.1rem 0.5rem;
  font-size: 0.85em;
  font-weight: 700;
}
</style>

# Contents {.unnumbered}

1. [Project overview](#project-overview)
2. [Project setup](#project-setup)
   - [Locate the repository](#locate-the-repository)
   - [Load the current workflow packages](#load-the-current-workflow-packages)
3. [miRNA input inspection and expression-source decision](#mirna-input-inspection-and-expression-source-decision)
   - [Inventory the raw miRNA CEL files](#inventory-the-raw-mirna-cel-files)
   - [Read the miRNA sample metadata](#read-the-mirna-sample-metadata)
   - [Confirm exact miRNA CEL-to-metadata matching](#confirm-exact-mirna-cel-to-metadata-matching)
   - [Read the miRNA CEL files](#read-the-mirna-cel-files)
   - [Validate miRNA raw intensities](#validate-mirna-raw-intensities)
   - [Review miRNA raw-intensity boxplots](#review-mirna-raw-intensity-boxplots)
   - [Review miRNA raw-intensity density curves](#review-mirna-raw-intensity-density-curves)
   - [Record the miRNA normalization limitation and expression source](#record-the-mirna-normalization-limitation-and-expression-source)
4. [Clariom S/mRNA sample preparation and normalization](#clariom-smrna-sample-preparation-and-normalization)
   - [Build and validate the mRNA sample sheet](#build-and-validate-the-mrna-sample-sheet)
   - [Read, inspect, and RMA-normalize the mRNA CEL files](#read-inspect-and-rma-normalize-the-mrna-cel-files)
   - [Validate the saved mRNA RMA output](#validate-the-saved-mrna-rma-output)
   - [Review the mRNA CEL and RMA figures](#review-the-mrna-cel-and-rma-figures)
5. [Clariom S Mouse HT biological annotation](#clariom-s-mouse-ht-biological-annotation)
   - [Annotation workflow and expected outputs](#annotation-workflow-and-expected-outputs)
   - [Install the biological annotation package](#install-the-biological-annotation-package)
   - [Inspect the annotation database](#inspect-the-annotation-database)
   - [Measure transcript-cluster mapping coverage](#measure-transcript-cluster-mapping-coverage)
   - [Build the validated annotation table](#build-the-validated-annotation-table)
   - [Prepare the primary analysis-ready mRNA dataset](#prepare-the-primary-analysis-ready-mrna-dataset)
6. [Analysis-ready mRNA sample quality control](#analysis-ready-mrna-sample-quality-control)
   - [Run sample-level quality control](#run-sample-level-quality-control)
   - [Review the quality-control outputs](#review-the-quality-control-outputs)
7. [Build the paired miRNA–mRNA sample manifest](#build-the-paired-mirnamrna-sample-manifest)
   - [Create and validate the paired manifest](#create-and-validate-the-paired-manifest)
   - [Review the manifest outputs](#review-the-manifest-outputs)
   - [Use the manifest for DE](#use-the-manifest-for-de)
8. [Exploratory mRNA DE](#exploratory-mrna-de)
   - [Define the limma models](#define-the-limma-models)
   - [Run the mRNA limma analysis](#run-the-mrna-limma-analysis)
   - [Review the mRNA DE results](#review-the-mrna-de-results)
9. [Multi-omics integration plan](#multi-omics-integration-plan)
   - [Obtain miRNA–mRNA target resources](#obtain-mirnamrna-target-resources)
   - [Download the validated miRTarBase interactions](#download-the-validated-mirtarbase-interactions)
   - [Download the predicted TargetScanMouse interactions](#download-the-predicted-targetscanmouse-interactions)
   - [Prepare and combine the target tables](#prepare-and-combine-the-target-tables)
   - [Identifier inspection and required corrections](#identifier-inspection-and-required-corrections)
   - [Inspect the differential-expression inputs](#inspect-the-differential-expression-inputs)
   - [Collapse evidence to unique measured pairs](#collapse-evidence-to-unique-measured-pairs)
   - [Review pairs passing the FDR threshold in both layers](#review-pairs-passing-the-fdr-threshold-in-both-layers)
   - [Test adjusted associations across paired animals](#test-adjusted-associations-across-paired-animals)
   - [Examine covariate effects on associations](#examine-covariate-effects-on-associations)
   - [Build the integrated pair-evidence summary](#build-the-integrated-pair-evidence-summary)
   - [Prepare enrichment inputs](#prepare-enrichment-inputs)
   - [Run mRNA GSEA](#run-mrna-gsea)
   - [Summarize and reduce GSEA redundancy](#summarize-and-reduce-gsea-redundancy)
   - [Run formal multi-omics ORA](#run-formal-multi-omics-ora)
10. [Record software versions](#record-software-versions)

<hr class="section-divider">

# Project overview

This project develops a reproducible R/Bioconductor pipeline for the
`mirna_150001` miRNA and `clariomS_150002` mRNA microarray experiments. The
two omics layers are processed separately because they use different array
platforms, platform-design packages, and expression-feature identifiers.

The documented workflow proceeds in this order:

```text
Set up the project environment
    ↓
Inspect and match miRNA inputs
    ↓
Document the usable miRNA expression source
    ↓
Build and validate the mRNA sample sheet
    ↓
Read, QC, and RMA-normalize the mRNA CEL files
    ↓
Annotate mRNA transcript clusters with biological gene identifiers
    ↓
Prepare the uniquely gene-mapped mRNA matrix
    ↓
Run sample-level QC on the analysis-ready mRNA matrix
    ↓
Match miRNA and mRNA samples by animal ID
    ↓
Freeze the paired analysis manifest
    ↓
Run exploratory paired-cohort differential expression (DE)
    ↓
Match external target evidence to exact measured features
    ↓
Create one row per measured miRNA–mRNA pair and contrast
```

Heavy or result-generating chunks are executable but use `eval: false` so that
rendering the notebook does not silently rerun analyses. Validation chunks read
the saved outputs and report their expected or observed results.

<hr class="section-divider">

# Project setup

## Locate the repository

This code allows the notebook to run from either the repository root or the
`notebooks/` directory. All later paths are constructed from `project_root`.

```{r}
#| label: locate-project

project_root <- if (file.exists("DESCRIPTION")) {
  normalizePath(".")
} else if (file.exists("../DESCRIPTION")) {
  normalizePath("..")
} else {
  stop("Run this notebook from the repository root or notebooks/ directory.")
}

project_root
```

**Expected outcome:** the absolute path ending in
`mirna-microarray-pipeline`.

## Load the current workflow packages

The project uses `renv` to restore and activate its R package environment. This
chunk loads that environment and checks the packages required by the workflow.

```{r}
#| label: load-packages

if (!requireNamespace("renv", quietly = TRUE)) {
  stop("Install renv before running this notebook.")
}

renv::load(project = project_root)

required_workflow_packages <- c(
  "Biobase", "oligo", "limma", "pd.mirna.4.1"
)
workflow_package_available <- vapply(
  required_workflow_packages,
  requireNamespace,
  quietly = TRUE,
  FUN.VALUE = logical(1)
)

workflow_package_available

if (!all(workflow_package_available)) {
  stop(
    "Missing required package(s): ",
    paste(
      names(workflow_package_available)[!workflow_package_available],
      collapse = ", "
    )
  )
}
```

**Expected outcome:** every required package check is `TRUE`.

<hr class="section-divider">

# miRNA input inspection and expression-source decision

## Inventory the raw miRNA CEL files

```{r}
#| label: list-cel-files

mirna_cel_directory <- file.path(
  project_root, "data/raw/mirna_150001"
)

mirna_cel_files <- list.files(
  mirna_cel_directory,
  pattern = "\\.[Cc][Ee][Ll]$",
  full.names = TRUE,
  recursive = TRUE
)

mirna_cel_files <- sort(mirna_cel_files)
mirna_cel_names <- basename(mirna_cel_files)

length(mirna_cel_files)
head(mirna_cel_names)
```

**Expected outcome:** one CEL file per expected array. A filename such as
`5505014461569112923888_A05_499422-50.CEL` contains an array prefix, an array
position, and a sample identifier. Matching below uses the complete filename,
not parsed filename components.

## Read the miRNA sample metadata

```{r}
#| label: read-metadata

mirna_metadata_file <- file.path(
  project_root,
  "metadata/Animal-Tracking-Sheet-RNA-R-miRNA.csv"
)

mirna_metadata <- read.csv(
  mirna_metadata_file,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

dim(mirna_metadata)
names(mirna_metadata)
head(mirna_metadata)
```

**Expected outcome:** one metadata row per expected array and columns including
`File Name`, `Animal ID`, `Sex`, `Age`, and `Treatment`.

## Confirm exact miRNA CEL-to-metadata matching

```{r}
#| label: match-cel-metadata

mirna_metadata_names <- mirna_metadata[["File Name"]]

mirna_matching_summary <- c(
  CEL_files = length(mirna_cel_names),
  metadata_rows = nrow(mirna_metadata),
  matched_CEL_files = sum(mirna_cel_names %in% mirna_metadata_names),
  CEL_without_metadata = sum(!mirna_cel_names %in% mirna_metadata_names),
  metadata_without_CEL = sum(!mirna_metadata_names %in% mirna_cel_names),
  duplicate_CEL_names = sum(duplicated(mirna_cel_names)),
  duplicate_metadata_names = sum(duplicated(mirna_metadata_names))
)

mirna_matching_summary

stopifnot(
  length(mirna_cel_names) == nrow(mirna_metadata),
  setequal(mirna_cel_names, mirna_metadata_names),
  !anyDuplicated(mirna_cel_names),
  !anyDuplicated(mirna_metadata_names)
)
```

**Expected outcome:** every CEL filename matches one metadata row, with no
missing, extra, or duplicated filenames.

## Read the miRNA CEL files

Each biological sample was hybridized to one physical array, producing one CEL
file. Consequently, the number of biological samples should equal the number
of biological arrays.

```{r}
#| label: read-cel-files

mirna_raw_cel <- oligo::read.celfiles(mirna_cel_files)
mirna_raw_cel
```

**Expected outcome:** an `ExpressionFeatureSet` containing every input array
and annotated with `pd.mirna.4.1`.

## Validate miRNA raw intensities

```{r}
#| label: validate-intensities

mirna_raw_intensities <- Biobase::exprs(mirna_raw_cel)

dim(mirna_raw_intensities)
table(is.finite(mirna_raw_intensities))
summary(as.vector(mirna_raw_intensities))
```

**Expected outcome:** a complete raw-intensity matrix with finite values, no
`NA` or `NaN` values, and the right-skewed distribution expected from
microarray probe intensities.

## Review miRNA raw-intensity boxplots

```{r}
#| label: raw-intensity-boxplots
#| fig-cap: "Raw log2 intensity distributions for the miRNA arrays."
#| fig-width: 12
#| fig-height: 7

boxplot(
  log2(mirna_raw_intensities),
  outline = FALSE,
  las = 2,
  names = sub("\\.[Cc][Ee][Ll]$", "", mirna_cel_names),
  main = "Raw CEL intensity distributions",
  ylab = "log2 raw intensity"
)
```

**Expected outcome:** broadly similar distributions with modest differences
between array medians and no visually extreme global-intensity outlier. This is
reassuring but is not sufficient by itself to declare every array high quality.

## Review miRNA raw-intensity density curves

```{r}
#| label: raw-intensity-densities
#| fig-cap: "Raw log2 intensity density curves for the miRNA arrays."
#| fig-width: 12
#| fig-height: 7

limma::plotDensities(
  log2(mirna_raw_intensities),
  legend = FALSE,
  main = "Raw CEL intensity densities",
  xlab = "log2 raw intensity"
)
```

**Expected outcome:** most curves overlap around the main peak. Differences in
peak shape and position indicate technical variation that would normally be
addressed through normalization and assessed with further QC. The long right
tail represents a small number of highly expressed probes.

## Record the miRNA normalization limitation and expression source

CEL-based normalization must produce a finite expression matrix before it is
used downstream. If that validation fails, use a separately validated
expression source and record the selected method in the analysis outputs.

**Expected outcome:** one validated miRNA expression source is selected before
annotation, QC, or DE.

<hr class="section-divider">

# Clariom S/mRNA sample preparation and normalization

The Clariom S experiment is analyzed separately from the miRNA experiment. Its
raw data are under `data/raw/clariomS_150002/`, and its generated outputs are
under `results/mrna/`.

## Build and validate the mRNA sample sheet

The first mRNA script inventories the CEL files, extracts the animal ID at the
start of each filename, and matches it exactly to the `Animal ID` column in
`metadata/Animal-Tracking-Sheet-RNA-Animal-ID.csv`.

```{r}
#| label: run-mrna-sample-sheet
#| eval: false

system2(
  "Rscript",
  file.path(
    project_root,
    "scripts/mrna/00_build_sample_sheet_mrna.R"
  )
)
```

This cell is executable manually. Automatic evaluation is disabled so that
rendering the notebook does not overwrite timestamped result summaries.

The script writes:

- `results/mrna/sample_sheet/sample_sheet_mrna_150002.csv`
- `results/mrna/sample_sheet/metadata_without_cel_mrna_150002.csv`
- `results/mrna/sample_sheet/sample_matching_summary_mrna_150002.txt`

```{r}
#| label: inspect-mrna-sample-sheet

mrna_sample_sheet_file <- file.path(
  project_root,
  "results/mrna/sample_sheet/sample_sheet_mrna_150002.csv"
)
mrna_matching_summary_file <- file.path(
  project_root,
  "results/mrna/sample_sheet/sample_matching_summary_mrna_150002.txt"
)

mrna_sample_sheet <- read.csv(
  mrna_sample_sheet_file,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

dim(mrna_sample_sheet)
table(mrna_sample_sheet$match_status)
table(
  mrna_sample_sheet$treatment[
    mrna_sample_sheet$match_status == "matched"
  ]
)
cat(readLines(mrna_matching_summary_file), sep = "\n")
```

**Expected outcome:** every biological CEL file matches one metadata row,
control arrays are identified, and matched filenames and animal IDs are
unique.

The validated metadata treatment labels are `Naive`, `Sham`, `MCAO1hr`,
`MCAO3hr`, and `MCAO24hr`. These names are used consistently across the mRNA,
miRNA, and paired-manifest workflows.

## Read, inspect, and RMA-normalize the mRNA CEL files

The second script reads all CEL files, verifies the raw intensities, creates
raw QC plots, runs RMA, validates the normalized expression matrix, and saves
the normalized data and post-RMA QC plots.

```{r}
#| label: run-mrna-cel-rma
#| eval: false

system2(
  "Rscript",
  file.path(
    project_root,
    "scripts/mrna/01_array_qc_normalize_mrna.R"
  )
)
```

This heavy cell is executable manually. Automatic evaluation is disabled
because it reads the CEL files and reruns RMA.

**Expected outcome:** the CEL headers resolve to the
`pd.clariom.s.mouse.ht` platform-design package.

## Validate the saved mRNA RMA output

```{r}
#| label: validate-mrna-rma-output

mrna_rma_eset_file <- file.path(
  project_root,
  "results/mrna/normalized/rma_expression_set_mrna_150002.rds"
)
mrna_rma_summary_file <- file.path(
  project_root,
  "results/mrna/qc/cel_rma/cel_rma_summary_mrna_150002.txt"
)

mrna_rma_eset <- readRDS(mrna_rma_eset_file)
mrna_rma_matrix <- Biobase::exprs(mrna_rma_eset)

dim(mrna_rma_matrix)
table(is.finite(mrna_rma_matrix))
summary(as.vector(mrna_rma_matrix))
cat(readLines(mrna_rma_summary_file), sep = "\n")
```

**Expected outcome:** the saved RMA matrix contains every normalized array,
uses transcript-cluster row identifiers, and contains no `NA`, `NaN`, or
infinite expression values.

## Review the mRNA CEL and RMA figures

```{r}
#| label: display-mrna-qc-figures
#| echo: true
#| fig-cap:
#|   - "Raw Clariom S CEL intensity distributions."
#|   - "Raw Clariom S CEL intensity density curves."
#|   - "RMA-normalized Clariom S expression distributions."
#|   - "RMA-normalized Clariom S expression density curves."

mrna_qc_dir <- file.path(
  project_root,
  "results/mrna/qc/cel_rma"
)

mrna_qc_figures <- file.path(
  mrna_qc_dir,
  c(
    "raw_cel_intensity_boxplot_mrna_150002.png",
    "raw_cel_intensity_density_mrna_150002.png",
    "rma_expression_boxplot_mrna_150002.png",
    "rma_expression_density_mrna_150002.png"
  )
)

stopifnot(all(file.exists(mrna_qc_figures)))
knitr::include_graphics(mrna_qc_figures)
```

**Expected outcome:** raw arrays may show technical variation, whereas
RMA-normalized biological arrays should have broadly comparable expression
distributions. Control arrays remain identified and are excluded from
biological contrasts.

<hr class="section-divider">

# Clariom S Mouse HT biological annotation

RMA produces expression values indexed by `transcript_cluster_id`.
`pd.clariom.s.mouse.ht` was required to summarize probes into those transcript
clusters, but it does not provide the gene symbols and stable gene identifiers
needed for biological interpretation. The next workflow stage therefore maps
the transcript clusters using the matching Clariom S Mouse HT biological
annotation package.

![Relationship among probes, probesets, transcript isoforms, transcript clusters, and gene annotation. The first panel shows the array measurement hierarchy, the second shows two Gene A isoforms represented by related probesets within one transcript cluster, and the third shows a transcript cluster that maps to more than one gene.](images/microarray_probes_transcript_clusters_gene_annotation_clear.png){#fig-mrna-transcript-cluster-concepts width="100%" fig-alt="Three-panel educational diagram. Panel one shows a biological sample measured on one microarray, with probes grouped into probesets and related probesets grouped into transcript clusters. Panel two shows Gene A producing two transcript isoforms, each measured by a probeset, with both probesets summarized into Transcript cluster A. Panel three shows an ambiguous transcript cluster containing two probesets and mapping to Gene 1 and Gene 2."}

## Annotation workflow and expected outputs

```text
RMA expression matrix
results/mrna/normalized/rma_expression_mrna_150002.csv
    ↓
Install and record clariomsmousehttranscriptcluster.db
    ↓
Inspect supported keys and biological annotation fields
    ↓
Map transcript_cluster_id values and measure mapping coverage
    ↓
Write validated annotation and annotated expression outputs
results/mrna/annotation/
    ↓
Keep biological samples and uniquely gene-mapped transcript clusters
results/mrna/analysis_ready/
```

The preceding RMA validation already checks matrix dimensions and finite
expression values, so those checks are not repeated here.

## Install the biological annotation package

The `clariomsmousehttranscriptcluster.db` package provides the biological
mapping from Clariom S Mouse HT transcript-cluster IDs to gene symbols, Entrez
IDs, gene names, and other identifiers. It complements
`pd.clariom.s.mouse.ht`, which was used for RMA normalization but does not
provide the gene-level interpretation required downstream.

```{r}
#| label: install-mrna-biological-annotation-package
#| eval: false

BiocManager::install(
  "clariomsmousehttranscriptcluster.db",
  ask = FALSE,
  update = FALSE
)

renv::record(
  c(
    "AnnotationDbi",
    "KEGGREST",
    "org.Mm.eg.db",
    "clariomsmousehttranscriptcluster.db"
  )
)
```

Automatic evaluation is disabled because package installation changes the
project environment. Run this cell once manually.

**Expected outcome:** the platform annotation package and its dependencies are
available in the project library and recorded in `renv.lock`.

## Inspect the annotation database

Before mapping the expression matrix, inspect the database object to confirm
its organism, platform, valid join key, and available biological fields. The
mRNA matrix identifier `transcript_cluster_id` will be matched to the database
key type `PROBEID`.

```{r}
#| label: inspect-mrna-biological-annotation

suppressPackageStartupMessages(library(AnnotationDbi))
suppressPackageStartupMessages(
  library(clariomsmousehttranscriptcluster.db)
)

mrna_annotation_db <- clariomsmousehttranscriptcluster.db

class(mrna_annotation_db)

mrna_required_keytype <- "PROBEID"
mrna_required_annotation_columns <- c(
  "SYMBOL",
  "ENTREZID",
  "ENSEMBL",
  "GENENAME",
  "GENETYPE",
  "REFSEQ",
  "UNIPROT"
)

mrna_required_keytype %in% AnnotationDbi::keytypes(mrna_annotation_db)
mrna_required_annotation_columns %in%
  AnnotationDbi::columns(mrna_annotation_db)
```

**Expected outcome:**

- the object is a `ChipDb` biological annotation database;
- the organism is `Mus musculus` (`TAXID` 10090);
- the chip name is `clariomsmouseht`;
- `PROBEID` is the appropriate transcript-cluster join key;
- useful output fields include `SYMBOL`, `ENTREZID`, `ENSEMBL`, `GENENAME`,
  `GENETYPE`, `REFSEQ`, and `UNIPROT`;
- database metadata reports the annotation source versions and dates needed
  for reproducibility.

## Measure transcript-cluster mapping coverage

This step checks whether the normalized matrix identifiers can be joined
reliably to the annotation database before any annotated table is written. It
compares the existing RMA row names with `PROBEID`, measures annotation
coverage, and flags one-to-many gene mappings that must not be silently
collapsed.

```{r}
#| label: measure-mrna-annotation-coverage

mrna_transcript_cluster_ids <- rownames(mrna_rma_matrix)
mrna_database_probe_ids <- AnnotationDbi::keys(
  mrna_annotation_db,
  keytype = "PROBEID"
)

mrna_matched_probe_ids <- intersect(
  mrna_transcript_cluster_ids,
  mrna_database_probe_ids
)
mrna_unmatched_probe_ids <- setdiff(
  mrna_transcript_cluster_ids,
  mrna_database_probe_ids
)

mrna_annotation_preview <- AnnotationDbi::select(
  mrna_annotation_db,
  keys = mrna_matched_probe_ids,
  keytype = "PROBEID",
  columns = c("SYMBOL", "ENTREZID")
)

mrna_symbols_per_probe <- tapply(
  mrna_annotation_preview$SYMBOL,
  mrna_annotation_preview$PROBEID,
  function(values) {
    length(unique(values[!is.na(values) & nzchar(values)]))
  }
)

mrna_mapping_summary <- c(
  expression_transcript_clusters = length(mrna_transcript_cluster_ids),
  duplicate_expression_ids = sum(duplicated(mrna_transcript_cluster_ids)),
  matched_probe_ids = length(mrna_matched_probe_ids),
  unmatched_probe_ids = length(mrna_unmatched_probe_ids),
  matched_with_gene_symbol = sum(mrna_symbols_per_probe > 0),
  probe_ids_with_multiple_symbols = sum(mrna_symbols_per_probe > 1)
)

mrna_mapping_summary
```

**Expected outcome:** most transcript-cluster IDs match the platform database,
expression IDs are unique, and unmatched or one-to-many mappings are counted
explicitly before filtering.

## Build the validated annotation table

This step creates the reusable transcript-cluster-to-gene annotation. The
script joins `transcript_cluster_id` to `PROBEID`, retains unmatched features
with an explicit status, and writes one row per gene mapping so that
one-to-many relationships are not silently reduced to the first match.

```{r}
#| label: build-mrna-transcript-cluster-annotation
#| eval: false

system2(
  "Rscript",
  file.path(
    project_root,
    "scripts/mrna/02_annotate_transcript_clusters_mrna.R"
  )
)
```

Automatic evaluation is disabled because the script writes result files. It
creates:

- `results/mrna/annotation/transcript_cluster_gene_annotation_mrna_150002.csv`
- `results/mrna/annotation/unmatched_transcript_clusters_mrna_150002.csv`
- `results/mrna/annotation/transcript_cluster_annotation_summary_mrna_150002.txt`

**Expected outcome:** every expression transcript cluster is represented in
the long-format table, unmatched features remain identifiable, one-to-many
gene mappings are preserved, and no exact duplicate annotation rows are
created.

## Prepare the primary analysis-ready mRNA dataset

The normalized RMA matrix still contains two control arrays and several feature
types that should not enter the primary gene-level analysis. This step creates
a cleaner dataset while leaving the original normalized matrix unchanged.

For samples, the script keeps biological arrays that match the animal metadata
and removes control arrays. For features, it keeps a transcript
cluster only when it is not an Affymetrix control and maps to exactly one gene.
It removes transcript clusters that are absent from the annotation database,
have no gene assignment, or map to more than one gene. Ambiguous mappings are
not discarded silently: every input transcript cluster and its filtering reason
are recorded in a separate report.

The retained expression rows and annotation rows are written in the same order.
This makes it possible to join them directly by `transcript_cluster_id` and
prevents an expression profile from being assigned to the wrong gene.

```{r}
#| label: prepare-analysis-ready-mrna
#| eval: false

system2(
  "Rscript",
  file.path(
    project_root,
    "scripts/mrna/03_prepare_analysis_ready_mrna.R"
  )
)
```

Automatic evaluation is disabled because the script writes result files.

The script creates the following outputs:

| Output | Contents |
|---|---|
| <span class="downstream-file">`expression_matrix_unique_gene_mapped_mrna.csv`</span> | RMA expression for retained transcript clusters across the biological samples. The first column is `transcript_cluster_id`. |
| <span class="downstream-file">`transcript_cluster_annotation_unique_gene_mapped_mrna.csv`</span> | One annotation row per retained transcript cluster, with Entrez ID, symbol, gene name, gene type, and Ensembl ID. |
| <span class="downstream-file">`analysis_samples_mrna.csv`</span> | Metadata for the biological samples represented by the expression columns, including animal ID, sex, age, treatment, and CEL filename. |
| `transcript_cluster_filter_report_mrna.csv` | Every original transcript cluster, whether it was retained, and the reason for any exclusion. |
| `analysis_ready_summary_mrna.txt` | Input paths, filtering rules, retained and excluded counts, matrix dimensions, and validation checks. |

All five files are written under `results/mrna/analysis_ready/`.

**Expected outcome:** expression and annotation contain the same uniquely
gene-mapped transcript clusters in identical order, sample columns match the
analysis metadata, and all retained expression values are finite.

<hr class="section-divider">

# Analysis-ready mRNA sample quality control

The analysis-ready expression matrix has passed feature annotation and
filtering, but the biological samples must also be evaluated before
DE. This stage checks normalized expression
distributions, major sources of sample variation, sample-to-sample similarity,
and unusually distant samples. It reports possible outliers for review only
and does not remove samples.

The QC script uses these two outputs from the preceding section:

- `results/mrna/analysis_ready/expression_matrix_unique_gene_mapped_mrna.csv`
- `results/mrna/analysis_ready/analysis_samples_mrna.csv`

## Run sample-level quality control

```{r}
#| label: run-analysis-ready-mrna-sample-qc
#| eval: false

system2(
  "Rscript",
  file.path(
    project_root,
    "scripts/mrna/04_sample_qc_mrna.R"
  )
)
```

Automatic evaluation is disabled because the script writes result files. The
script verifies sample alignment and finite expression values, then calculates
expression summaries, PCA, hierarchical clustering, distance from the
expression centroid, median inter-sample correlation, and distance in the
first two principal components.

A sample is flagged for review if it is more than three standard deviations
from the cohort for high expression distance, high PCA distance, or low median
sample correlation. These flags are screening aids, not automatic exclusion
decisions.

## Review the quality-control outputs

The script writes all outputs under `results/mrna/qc/analysis_ready/`:

| Output | Contents |
|---|---|
| `expression_boxplot_mrna.png` and `expression_density_mrna.png` | Normalized expression distributions across the analysis-ready samples. |
| `pca_by_treatment_mrna.png`, `pca_by_sex_mrna.png`, `pca_by_age_group_mrna.png`, and `pca_by_array_prefix_mrna.png` | The same PCA coordinates labelled by key biological and technical variables. |
| `sample_clustering_mrna.png` | Unsupervised hierarchical clustering of samples using the analysis-ready expression matrix. |
| `pca_scores_mrna.csv` | Sample metadata, PC1 and PC2 scores, explained variance, and two-dimensional PCA distance. |
| `sample_qc_metrics_mrna.csv` | Per-sample expression summaries, distance and correlation metrics, and all review flags. |
| `possible_outliers_mrna.csv` | Only samples that meet at least one predefined review rule; the file may contain zero rows. |
| `sample_qc_summary_mrna.txt` | Input paths, dimensions, finite-value checks, explained variance, flag rules, and the number of samples flagged. |

**Expected outcome:** the QC matrix contains only finite values, expression
distributions are broadly comparable, PCA and clustering files are created,
and any samples meeting a predefined review rule are reported without
automatic exclusion.

<hr class="section-divider">

# Build the paired miRNA–mRNA sample manifest

The molecular layers must be aligned explicitly before any paired multi-omics
analysis. This step matches the selected `rma_normalized_mirna` samples to the
analysis-ready mRNA samples by exact `animal_id`, checks sex and treatment
agreement, and confirms that both corresponding filenames occur in their
expression matrices.

The analysis manifest keeps the treatment groups defined for the paired
analysis. Samples outside those groups are recorded in the run summary but are
not written to a second manifest.

## Create and validate the paired manifest

```{r}
#| label: build-paired-mirna-mrna-manifest
#| eval: false

system2(
  "Rscript",
  file.path(
    project_root,
    "scripts/multiomics/00_build_paired_sample_manifest.R"
  )
)
```

Automatic evaluation is disabled because the script writes result files. The
miRNA and mRNA metadata already use the standardized labels `Sham`, `MCAO1hr`,
`MCAO3hr`, and `MCAO24hr`. The script requires treatment and sex to agree
across layers for every pair and records MD5 checksums of all four input files.

## Review the manifest outputs

All outputs are written under `results/multiomics/sample_manifest/`:

| Output | Contents |
|---|---|
| `paired_manifest_mirna_mrna.csv` | One row per retained animal, with standardized treatment and the matching miRNA and mRNA filenames. |
| `unpaired_samples_mirna_mrna.csv` | Biological samples present in only one molecular layer, with the missing layer recorded. |
| `paired_manifest_summary.txt` | Input files and checksums, pair counts, treatment counts, exclusions, and validation results. |

**Expected outcome:** every retained animal has one miRNA file and one mRNA
file, animal IDs are unique, sex and standardized treatment agree across
layers, both filenames exist in their selected expression matrices, and
unpaired or excluded samples are reported separately.

## Use the manifest for DE

The paired manifest is the single source of sample eligibility for paired
miRNA and mRNA analyses. Large expression matrices are not copied into new
paired-sample files. Instead, each DE script reads
`paired_manifest_mirna_mrna.csv`, selects the corresponding expression columns
in memory, and preserves the manifest animal order.

The existing miRNA limma script applies this rule and supports
`treatment_only`, additive `treatment_sex`, or additive `treatment_age`
modeling, as well as the combined additive `treatment_sex_age` model. Every
mode tests MCAO1hr versus Sham, MCAO3hr versus Sham, and MCAO3hr versus
MCAO1hr. Each limma summary records the selected model, manifest path,
included sample count, and any paired animals unavailable in the selected
expression dataset.

```bash
Rscript scripts/mirna/05_differential_expression_mirna.R \
  <dataset> <treatment_only|treatment_sex|treatment_age|treatment_sex_age>
```

**Expected outcome:** the selected design matrix is full rank, contains the
treatment coefficients and the requested adjustment coefficient, and produces
all three treatment-contrast result tables in a model-specific output folder.

<hr class="section-divider">

# Exploratory mRNA DE

The analysis-ready mRNA matrix has passed normalization, biological
annotation, feature filtering, sample-level QC, and paired-animal selection.
The next step estimates treatment-associated expression differences across
the mice in `paired_manifest_mirna_mrna.csv`. Expression columns are selected
and ordered in memory, so no duplicate paired-sample expression matrix is
written.

## Define the limma models

The script supports four model choices:

| Model argument | Formula |
|---|---|
| `treatment_only` | `~ 0 + treatment` |
| `treatment_sex` | `~ 0 + treatment + sex` |
| `treatment_age` | `~ 0 + treatment + age_group` |
| `treatment_sex_age` | `~ 0 + treatment + sex + age_group` |

Every model uses the paired Sham, MCAO1hr, and MCAO3hr groups and tests:

| Contrast | Limma definition | Positive logFC means |
|---|---|---|
| `MCAO1hr_vs_Sham` | `MCAO1hr - Sham` | Higher expression in MCAO1hr |
| `MCAO3hr_vs_Sham` | `MCAO3hr - Sham` | Higher expression in MCAO3hr |
| `MCAO3hr_vs_MCAO1hr` | `MCAO3hr - MCAO1hr` | Higher expression in MCAO3hr |

## Run the mRNA limma analysis

```{r}
#| label: run-mrna-treatment-only-limma
#| eval: false

system2(
  "Rscript",
  c(
    file.path(
      project_root,
      "scripts/mrna/05_differential_expression_mrna.R"
    ),
    "<treatment_only|treatment_sex|treatment_age|treatment_sex_age>"
  )
)
```

The script validates the manifest samples, preserves their order, fits the
three contrasts with `limma`, and joins gene annotation to each result. If the
model argument is omitted, `treatment_only` is used.

**Expected outcome:** the selected design matrix is full rank, contains the
treatment coefficients and requested adjustment coefficients, and writes its
results without changing other model folders.

## Review the mRNA DE results

The script writes all outputs under
`results/mrna/differential_expression/<model>_limma/`:

| Output | Contents |
|---|---|
| `analysis_samples_mrna.csv` | The analyzed mice in paired-manifest order, with their validated metadata. |
| `design_matrix_mrna.csv` | The treatment-only design matrix. |
| `contrast_matrix_mrna.csv` | The three contrast definitions and their coefficient directions. |
| `transcript_cluster_<contrast>_mrna.csv` | Limma statistics for every retained transcript cluster, with Entrez ID, symbol, gene name, gene type, and Ensembl ID. |
| `limma_summary_mrna.txt` | Dataset, model, manifest, sample counts, transcript-cluster count, contrasts, and raw/BH-adjusted result counts. |

**Expected outcome:** samples appear in exact manifest order, treatment counts
match the manifest, every result table contains one row per retained transcript
cluster, annotations are complete, and `logFC`, p-values, and BH-adjusted
p-values are finite.

<hr class="section-divider">

# Multi-omics integration plan

The project contains miRNA and mRNA measurements from the same 43 mice, so
each miRNA profile can be aligned with the mRNA profile from the same animal.
We will first identify [treatment-associated miRNAs](#use-the-manifest-for-de)
and [treatment-associated genes](#review-the-mrna-de-results) separately. We
will then ask whether a treatment-associated miRNA is linked to a documented
or predicted mouse target gene and whether their expression levels show the
expected relationship across the paired animals. For a repressive miRNA, for
example, higher miRNA expression together with lower target-gene expression
would be a compatible pattern. This pattern is supportive evidence only; it
does not prove direct regulation or causality.

<div class="purpose-box">

The primary integration question is:

> Which treatment-associated miRNAs are linked to treatment-associated mouse
> genes through documented or predicted target relationships, and do those
> miRNA–gene pairs show a compatible expression relationship in the same
> animals?

</div>

This is a **late-integration** strategy. The miRNA and mRNA datasets remain
separate during normalization, QC, annotation, and DE.
Their results are connected only afterward using:

1. **miRNA treatment effects:** whether a miRNA’s expression differs between
   the defined treatment groups.
2. **mRNA treatment effects:** whether a target gene’s expression differs
   between those same treatment groups.
3. **miRNA–gene target evidence:** whether an external resource documents or
   predicts that the miRNA can target that mouse gene.
4. **Paired-animal expression relationships:** whether the miRNA and target
   gene levels are related across the same mice, after accounting for the
   relevant design variables.
5. **Consistency across model choices:** whether the direction and strength
   of the evidence are reasonably similar in the treatment-only and adjusted
   models, rather than appearing in only one specification.

The primary expression model is:

```text
expression ~ treatment + sex + age
```

The treatment contrasts are:

```text
MCAO1hr vs Sham
MCAO3hr vs Sham
MCAO3hr vs MCAO1hr
```

Treatment-only, treatment-plus-sex, and treatment-plus-age models are used as
sensitivity analyses. The analysis does not estimate mediation or prove that
a miRNA causes an mRNA change because no stroke-severity outcome or temporal
measurement is available.

### Starting point

The first integration step is to obtain versioned mouse miRNA–target resources:

- **miRTarBase 10.0:** experimentally curated mouse miRNA–gene interactions;
- **TargetScanMouse 8.0:** computationally predicted mouse miRNA targets.

These files are needed to create the lookup relationship:

```text
measured mouse miRNA → candidate mouse target gene
```

They will then be matched to:

- the [annotated miRNA DE results](../results/mirna/differential_expression/);
- the [uniquely gene-mapped mRNA DE results](../results/mrna/differential_expression/);
- the [43-mouse paired manifest](../results/multiomics/sample_manifest/paired_manifest_mirna_mrna.csv).

The initial integration inputs are:

```text
results/multiomics/sample_manifest/paired_manifest_mirna_mrna.csv
results/mirna/differential_expression/
results/mrna/differential_expression/
resources/mirna_target_databases/mirtarbase_10.0/mmu_MTI.csv
resources/mirna_target_databases/targetscanmouse_8.0/
```

The current outputs include an evidence-level table, a unique measured-pair
table, an FDR candidate table, adjusted paired-animal association results, a
comparison of covariate-adjustment models, and an integrated pair-evidence
summary. Gene rankings, gene lists, backgrounds, and miRNA target sets are
also prepared for enrichment testing. Full-ranked mRNA GSEA and formal ORA
have now been run as separate analyses:

<div class="workflow-box">

```text
Paired manifest
      ↓
Separate miRNA and mRNA limma results
      ↓
Mouse miRNA–target resources
      ↓
Target-supported pair matching
      ↓
Exact-feature DE result joining
      ↓
Unique measured-pair evidence summary
      ↓
FDR candidate table
      ↓
Paired expression association
      ↓
Integrated miRNA–mRNA pair evidence
      ↓
Gene rankings, selected lists, and measured backgrounds
      ├── Full-ranked mRNA GSEA
      │         ↓
      │   Redundancy-reduced GSEA summary
      └── Formal 53-gene ORA with a 1,106-gene background
```

</div>

## Obtain miRNA–mRNA target resources

Target-supported integration requires a versioned lookup table connecting
mouse miRNAs to mouse target genes. These resources are external reference
data; they are not required for array normalization, sample QC, or limma.
The files are downloaded locally because the repository does not track large
external databases.

### Download the validated miRTarBase interactions

miRTarBase provides experimentally curated miRNA–target interactions. Open the
[current miRTarBase downloads page](https://awi.cuhk.edu.cn/miRTarBase/downloads/)
and download the **Mus musculus** species-specific file named `mmu_MTI.csv`.
Save it as:

```text
resources/mirna_target_databases/mirtarbase_10.0/mmu_MTI.csv
```

The current direct download is:

```text
https://awi.cuhk.edu.cn/miRTarBase/downloads/files/10.0/mmu_MTI.csv
```

The command-line equivalent is:

```bash
mkdir -p resources/mirna_target_databases/mirtarbase_10.0
curl -L --fail \
  https://awi.cuhk.edu.cn/miRTarBase/downloads/files/10.0/mmu_MTI.csv \
  -o resources/mirna_target_databases/mirtarbase_10.0/mmu_MTI.csv
```

Keep the original filename and record the release, download date, and SHA-256
checksum in a local source-metadata file. The table should contain mouse
miRNA identifiers, mouse target genes or Entrez IDs, experimental methods,
support type, and PubMed references.

### Download the predicted TargetScanMouse interactions

TargetScanMouse provides computationally predicted mouse miRNA targets. Use
the [TargetScanMouse 8.0 download page](https://www.targetscan.org/cgi-bin/targetscan/data_download.mmu80.cgi)
and download these files:

| File | Purpose |
|---|---|
| `miR_Family_Info.txt.zip` | miRNA-family, miRBase identifier, and sequence information |
| `Gene_info.txt.zip` | Mouse transcript, gene, and gene-symbol annotations |
| `Summary_Counts.default_predictions.txt.zip` | Default conserved mouse target predictions and scores |

Save and extract them under:

```text
resources/mirna_target_databases/targetscanmouse_8.0/
```

For example:

```bash
mkdir -p resources/mirna_target_databases/targetscanmouse_8.0
curl -L --fail \
  https://www.targetscan.org/mmu_80/mmu_80_data_download/miR_Family_Info.txt.zip \
  -o resources/mirna_target_databases/targetscanmouse_8.0/miR_Family_Info.txt.zip
curl -L --fail \
  https://www.targetscan.org/mmu_80/mmu_80_data_download/Gene_info.txt.zip \
  -o resources/mirna_target_databases/targetscanmouse_8.0/Gene_info.txt.zip
curl -L --fail \
  https://www.targetscan.org/mmu_80/mmu_80_data_download/Summary_Counts.default_predictions.txt.zip \
  -o resources/mirna_target_databases/targetscanmouse_8.0/Summary_Counts.default_predictions.txt.zip
unzip -o resources/mirna_target_databases/targetscanmouse_8.0/miR_Family_Info.txt.zip \
  -d resources/mirna_target_databases/targetscanmouse_8.0/
unzip -o resources/mirna_target_databases/targetscanmouse_8.0/Gene_info.txt.zip \
  -d resources/mirna_target_databases/targetscanmouse_8.0/
unzip -o resources/mirna_target_databases/targetscanmouse_8.0/Summary_Counts.default_predictions.txt.zip \
  -d resources/mirna_target_databases/targetscanmouse_8.0/
```

Filter TargetScan records to `Species ID = 10090` before matching them to the
mouse expression results. Preserve the original ZIP archives, extracted files,
release information, download date, and checksums. Both resource folders are
ignored by Git and should remain local.

### Prepare and combine the target tables

The two databases are prepared separately because their column names,
identifier formats, and evidence fields differ. The first script standardizes
the experimentally curated miRTarBase table. The second standardizes the
TargetScanMouse prediction table and retains its prediction scores. The third
combines both standardized tables while preserving the evidence source for
each relationship.

The miRTarBase script keeps only rows with mouse species fields and a
normalized `mmu-` miRNA identifier, excluding cross-species entries.

Run the scripts in this order from the repository root:

```bash
Rscript scripts/multiomics/01_prepare_mirtarbase_targets.R
Rscript scripts/multiomics/02_prepare_targetscanmouse_targets.R
Rscript scripts/multiomics/03_combine_mirna_target_evidence.R
Rscript scripts/multiomics/04_build_measured_mirna_mrna_target_pairs.R
Rscript scripts/multiomics/05_join_target_pairs_to_de_results.R
Rscript scripts/multiomics/06_summarize_target_pair_de_results.R
Rscript scripts/multiomics/07_collapse_target_evidence_to_unique_pairs.R
```

The scripts write these outputs under
`results/multiomics/miRNA_target_evidence/`:

| Output | Contents |
|---|---|
| `mirtarbase/mirtarbase_mouse_targets_standardized.csv` | One standardized row per distinct mouse miRNA–gene relationship from miRTarBase, including experimental evidence and PubMed references. |
| `targetscanmouse/targetscanmouse_targets_standardized.csv` | Mouse TargetScanMouse relationships with representative miRNA, target gene, transcript, and prediction scores. |
| `combined/mirna_mouse_target_evidence.csv` | Combined evidence table retaining the source database for each relationship and indicating when a pair appears in both resources. |
| `analysis_ready/mirna_mrna_target_pairs.csv` | Target-supported relationships whose miRNAs and mRNA genes are present in the analysis-ready annotations, with the identifier used for each match. |
| `analysis_ready/mirna_mrna_de_supported_target_pairs.csv` | Long-format target-pair table joined to the three treatment–sex–age DABG20 miRNA and mRNA DE contrasts, including both layers' DE statistics. |
| `analysis_ready/mirna_mrna_de_pair_evidence_summary.csv` | Evidence-row table with FDR < 0.10 flags, candidate categories, and inverse-logFC direction flags for each contrast. Separate database records and publications remain separate rows. |
| `analysis_ready/mirna_mrna_unique_pair_summary.csv` | One row per exact miRNA probeset–mRNA transcript cluster–contrast combination, with miRTarBase and TargetScanMouse evidence summarized in separate columns. |
| `analysis_ready/mirna_mrna_fdr_candidates.csv` | The unique measured pair–contrast rows for which both the miRNA and mRNA adjusted p-values are below 0.10. |

Script 04 normalizes miRNA IDs case-insensitively, strips version suffixes from
Ensembl IDs, and matches target genes to the measured mRNA annotation by Entrez
ID, Ensembl gene ID, or gene symbol. Script 05 then joins the three DABG20
miRNA and mRNA DE contrasts using the exact miRNA probeset and exact mRNA
transcript-cluster IDs already established in script 04. Gene identifiers are
used to validate that exact feature join, not to rematch DE results across
multiple transcript clusters belonging to the same gene.

Script 06 adds transparent FDR and direction categories while retaining the
individual evidence records. Script 07 collapses those records to one row per
exact measured pair and contrast. It keeps miRTarBase publication and
experimental evidence separate from TargetScanMouse transcript and prediction
evidence rather than treating the two resources as equivalent. Script 07 also
writes the subset passing the FDR < 0.10 threshold in both molecular layers to
`mirna_mrna_fdr_candidates.csv`.

### Identifier inspection and required corrections

The standardized tables were inspected before target-pair matching. The main
observations were:

- miRTarBase miRNA IDs are lowercase (`mmu-mir-...`), whereas the miRNA
  annotation uses mixed case (`mmu-miR-...`).
- TargetScanMouse miRNA IDs are largely compatible after case normalization.
- miRTarBase target Entrez IDs overlap the mRNA annotation.
- TargetScanMouse's `target_entrez_id` field contains Ensembl gene IDs, not
  Entrez IDs.

The required corrections were implemented in the preparation and matching
scripts:

→ miRNA identifiers are converted to a common lowercase format before joining.

→ TargetScanMouse's former `target_entrez_id` field is now represented as the
original gene ID plus a corrected Ensembl gene-ID field; version suffixes such
as `.7` are removed for matching.

→ Target genes are matched to the mRNA annotation using Entrez IDs, Ensembl
gene IDs, or normalized gene symbols, while retaining the original identifiers
and evidence source.

### Inspect the differential-expression inputs

Before matching target relationships to differential-expression results, list
the available files and confirm their identifier columns:

```{r}
list.files(
  c(
    "results/mirna/differential_expression/rma_normalized_mirna/treatment_sex_age_limma",
    "results/mrna/differential_expression/treatment_sex_age_limma"
  ),
  full.names = TRUE
)
```

The inspection confirms that both datasets contain all three treatment
contrasts and use the treatment–sex–age model. miRNA results identify features
with `ProbeSetName` and miRNA names with `Transcript ID(Array Design)`.
mRNA results identify features with `transcript_cluster_id` and genes with
`ENTREZID`, `SYMBOL`, and `ENSEMBL`.

### Collapse evidence to unique measured pairs

The DE-supported table intentionally retains separate rows for distinct
miRTarBase publications, experimental methods, and TargetScanMouse prediction
records. Those rows are useful for evidence provenance but should not be
counted as independent miRNA–mRNA candidates.

Run the evidence-collapsing script from the repository root:

```bash
Rscript scripts/multiomics/07_collapse_target_evidence_to_unique_pairs.R
```

The unique key is:

```text
miRNA probeset ID + mRNA transcript-cluster ID + treatment contrast
```

The script first verifies that all DE statistics and candidate flags are
constant within that exact key. It then writes the complete unique-pair table
and the smaller FDR candidate table:

```text
results/multiomics/miRNA_target_evidence/analysis_ready/
  mirna_mrna_unique_pair_summary.csv
  mirna_mrna_fdr_candidates.csv
```

The output preserves separate support indicators, record counts, identifiers,
references, experimental evidence, transcripts, and prediction-score summaries
for miRTarBase and TargetScanMouse. It does not average transcript clusters,
select the smallest p-value, run paired-animal association tests, or perform
biological interpretation.

**Expected outcome:** each measured miRNA probeset–mRNA transcript
cluster–contrast key appears exactly once, all evidence rows are accounted for,
and the two target resources remain distinguishable.

### Review pairs passing the FDR threshold in both layers

Script 06 labels a row `both_layers_fdr_lt_0.10` when the adjusted p-value is
below 0.10 for both the miRNA and its matched mRNA transcript cluster in the
same treatment contrast. Before evidence records were collapsed, 59 rows met
this rule. Some of those rows described the same measured pair more than once
because they came from separate target-database records or publications.

Script 07 groups rows using the exact miRNA probeset, exact mRNA transcript
cluster, and treatment contrast. After that grouping, the 59 evidence rows
represent 56 distinct measured pair–contrast rows. All 56 occur in the
`MCAO3hr_vs_MCAO1hr` contrast.

In plain language, these 56 rows describe links between 6 miRNAs and 53 genes.
Every gene is linked to at least one of the six miRNAs. The row count is larger
than the gene count because `Asxl2`, `Mmp16`, and `Prkcb` are each linked to
two miRNAs. None of the 53 genes is represented by multiple mRNA transcript
clusters within this candidate set.

Thirteen pairs have inverse effect directions: one layer increases while the
other decreases. The other 43 pairs change in the same direction. Target
support consists of 21 miRTarBase-only pairs, 32 TargetScanMouse-only pairs,
and 3 pairs supported by both resources. These categories describe supporting
evidence; they do not prove that a miRNA directly regulates a gene.

For subsequent review or association analysis, read both new outputs and
confirm that the saved candidate table exactly matches the documented filter:

```{r}
#| label: inspect-fdr-candidate-pairs
#| eval: false

unique_pair_file <- file.path(
  project_root,
  "results/multiomics/miRNA_target_evidence/analysis_ready",
  "mirna_mrna_unique_pair_summary.csv"
)
fdr_candidate_file <- file.path(
  project_root,
  "results/multiomics/miRNA_target_evidence/analysis_ready",
  "mirna_mrna_fdr_candidates.csv"
)

unique_pair_summary <- read.csv(
  unique_pair_file,
  stringsAsFactors = FALSE,
  check.names = FALSE
)
fdr_candidates <- read.csv(
  fdr_candidate_file,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

expected_fdr_candidates <- unique_pair_summary[
  unique_pair_summary$mirna_fdr_lt_0_10 &
    unique_pair_summary$mrna_fdr_lt_0_10,
  ,
  drop = FALSE
]

pair_key <- function(data) {
  paste(
    data$mirna_probeset_id,
    data$transcript_cluster_id,
    data$contrast,
    sep = "||"
  )
}

c(
  measured_pairs = nrow(fdr_candidates),
  unique_mirnas = length(unique(fdr_candidates$mirna_id)),
  unique_genes = length(unique(fdr_candidates$gene_entrez_id)),
  inverse_pairs = sum(fdr_candidates$inverse_logFC_direction)
)
table(fdr_candidates$contrast)
table(
  miRTarBase = fdr_candidates$mirtarbase_supported,
  TargetScanMouse = fdr_candidates$targetscanmouse_supported
)

stopifnot(
  nrow(fdr_candidates) == 56L,
  !anyDuplicated(pair_key(fdr_candidates)),
  setequal(
    pair_key(fdr_candidates),
    pair_key(expected_fdr_candidates)
  ),
  all(fdr_candidates$mirna_fdr_lt_0_10),
  all(fdr_candidates$mrna_fdr_lt_0_10),
  all(
    fdr_candidates$contrast ==
      "MCAO3hr_vs_MCAO1hr"
  )
)
```

**Expected outcome:** the saved candidate table contains 56 non-duplicated
rows, exactly matches the filter applied to the complete unique-pair table,
and places all candidates in `MCAO3hr_vs_MCAO1hr`.

### Test adjusted associations across paired animals

Association asks whether animals with higher expression of a particular miRNA
also tend to have higher or lower expression of its target mRNA. A negative
association means that higher miRNA expression tends to occur with lower
target-mRNA expression, which is compatible with possible miRNA repression.
A positive association means that the miRNA and mRNA tend to change together.
Neither result proves direct regulation or causality.

Script 08 retrieves the miRNA and mRNA expression values for each of the 56
candidate pairs in the same 43 mice. It aligns the two molecular layers using
the frozen paired manifest and verifies that animals and measured features are
complete, unique, and in the same order.

For each pair, the adjusted model is:

```text
mRNA expression ~ miRNA expression + treatment + sex + age
```

The miRNA coefficient measures the expression relationship after accounting
for treatment, sex, and age. This adjustment is important because a miRNA and
mRNA could otherwise appear related simply because both differ between
treatment groups or by sex or age. The script also reports the unadjusted
Pearson correlation for reference, but the adjusted miRNA coefficient is the
primary association test. Benjamini–Hochberg correction is applied across the
56 pair-level p-values.

Run the association script from the repository root:

```bash
Rscript scripts/multiomics/08_test_adjusted_paired_associations.R
```

> **Script 08 outputs**
>
> - `results/multiomics/paired_association/mirna_mrna_adjusted_association_results.csv`
>   contains one row per tested measured pair. It retains the DE results and
>   target-source evidence and adds the unadjusted correlation, adjusted
>   slope, partial correlation, confidence interval, p-value, and pair-level
>   FDR.
>
> - `results/multiomics/paired_association/mirna_mrna_adjusted_association_summary.txt`
>   records the sample counts, model, number of tests, association-direction
>   counts, significance counts, and the pair with the smallest adjusted
>   association p-value.

Inspect and validate the saved results:

```{r}
#| label: inspect-adjusted-paired-associations
#| eval: false

association_file <- file.path(
  project_root,
  "results/multiomics/paired_association",
  "mirna_mrna_adjusted_association_results.csv"
)
association_results <- read.csv(
  association_file,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

association_pair_key <- paste(
  association_results$mirna_probeset_id,
  association_results$transcript_cluster_id,
  sep = "||"
)

c(
  tested_pairs = nrow(association_results),
  paired_animals = unique(association_results$n_paired_animals),
  negative_associations = sum(
    association_results$negative_adjusted_association
  ),
  positive_associations = sum(
    !association_results$negative_adjusted_association
  ),
  raw_p_lt_0_05 = sum(
    association_results$adjusted_P.Value < 0.05
  ),
  association_fdr_lt_0_10 = sum(
    association_results$adjusted_fdr_lt_0_10
  ),
  association_fdr_lt_0_05 = sum(
    association_results$adjusted_fdr_lt_0_05
  )
)

head(
  association_results[
    ,
    c(
      "association_rank", "mirna_id", "gene_symbol",
      "adjusted_beta", "adjusted_partial_r",
      "adjusted_P.Value", "adjusted_adj.P.Val",
      "association_direction"
    )
  ],
  10L
)

stopifnot(
  nrow(association_results) == 56L,
  !anyDuplicated(association_pair_key),
  all(association_results$n_paired_animals == 43L),
  all(is.finite(association_results$adjusted_beta)),
  all(is.finite(association_results$adjusted_P.Value)),
  isTRUE(all.equal(
    association_results$adjusted_adj.P.Val,
    p.adjust(association_results$adjusted_P.Value, method = "BH")
  ))
)
```

**Expected outcome after running Script 08 on the 56 selected candidate
pairs:** all 56 pairs are tested across all 43 paired animals and produce
can becomplete finite statistics. Nineteen adjusted associations are negative and
37 are positive. No adjusted association has a raw p-value below 0.05, and
none passes pair-level FDR < 0.10 or FDR < 0.05. The smallest adjusted
association p-value is 0.0527 for `mmu-mir-223-3p`–`Fbxo8`, with pair-level
FDR 0.826. Therefore, these treatment-associated, target-supported pairs do
not show statistically supported expression association after adjustment for
treatment, sex, and age.

### Examine covariate effects on associations

The unadjusted results and the fully adjusted results answer different
questions. The unadjusted correlation asks whether the miRNA and mRNA move
together across all mice without considering why. The fully adjusted model
asks whether that relationship remains after accounting for treatment, sex,
and age.

This distinction is especially important here because the 56 pairs were
selected after both molecular layers showed DE in
`MCAO3hr_vs_MCAO1hr`. A miRNA and mRNA can therefore appear correlated across
all mice simply because both respond to treatment. An unadjusted correlation
in this selected set is descriptive and is not independent evidence of
miRNA-mediated regulation.

Script 09 compares five pre-specified models:

1. no covariate adjustment across all 43 animals;
2. treatment adjustment across all 43 animals;
3. sex and age adjustment across all 43 animals;
4. treatment, sex, and age adjustment across all 43 animals; and
5. treatment, sex, and age adjustment using only the 27
   `MCAO1hr` and `MCAO3hr` animals.

The fifth model focuses on the two groups that generated the 56 candidates.
It is a supporting analysis and does not replace the primary 43-animal model.
Benjamini–Hochberg correction is applied separately across the 56 pair tests
within each model.

Run the covariate-effects script from the repository root:

```bash
Rscript scripts/multiomics/09_examine_covariate_effects_on_associations.R
```

> **Script 09 outputs**
>
> - `results/multiomics/paired_association/covariate_effects/mirna_mrna_association_model_comparison_long.csv`
>   contains one row per measured pair per model, giving 280 rows for 56 pairs
>   across five models. It records the association slope, partial correlation,
>   confidence interval, raw p-value, FDR, direction, and sample count.
>
> - `results/multiomics/paired_association/covariate_effects/mirna_mrna_association_model_comparison_wide.csv`
>   places the five model results beside one another for each of the 56 pairs.
>   It also flags which unadjusted raw-p findings no longer have raw p < 0.05
>   after treatment, sex-and-age, or full adjustment.
>
> - `results/multiomics/paired_association/covariate_effects/mirna_mrna_association_model_comparison_summary.txt`
>   records the number of animals, raw-p findings, FDR findings, and
>   smallest-p pair for each model.
>
> - `results/multiomics/paired_association/covariate_effects/figures/`
>   contains treatment-colored plots for the ten smallest unadjusted p-values,
>   the 13 inverse-DE pairs, and a direct comparison of unadjusted versus fully
>   adjusted p-values.

Inspect and validate the saved model comparisons:

```{r}
#| label: inspect-covariate-effects-on-associations
#| eval: false

comparison_dir <- file.path(
  project_root,
  "results/multiomics/paired_association/covariate_effects"
)
comparison_long <- read.csv(
  file.path(
    comparison_dir,
    "mirna_mrna_association_model_comparison_long.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
comparison_wide <- read.csv(
  file.path(
    comparison_dir,
    "mirna_mrna_association_model_comparison_wide.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

model_counts <- do.call(
  rbind,
  lapply(
    split(comparison_long, comparison_long$model_id),
    function(data) {
      data.frame(
        animals = unique(data$n_animals),
        tested_pairs = nrow(data),
        raw_p_lt_0_05 = sum(data$raw_p_lt_0_05),
        fdr_lt_0_10 = sum(data$fdr_lt_0_10),
        fdr_lt_0_05 = sum(data$fdr_lt_0_05)
      )
    }
  )
)
model_counts

stopifnot(
  nrow(comparison_long) == 280L,
  nrow(comparison_wide) == 56L,
  all(table(comparison_long$model_id) == 56L),
  all(is.finite(comparison_long$beta)),
  all(is.finite(comparison_long$P.Value)),
  all(is.finite(comparison_long$adj.P.Val))
)
```

**Expected outcome after comparing the five association models:** the
unadjusted analysis has 32 pairs with raw p < 0.05. After treatment adjustment,
only six retain raw p < 0.05 and none passes FDR < 0.10. Adjustment for sex
and age without treatment leaves 28 raw-p findings, whereas the full
treatment, sex, and age model leaves none. In the 27-animal MCAO-focused
analysis, one pair has raw p < 0.05, but none passes FDR < 0.10.

These results show that treatment explains most of the apparent unadjusted
relationships, with sex and age accounting for additional variation. The
candidate pairs show coordinated treatment responses, but no pair currently
has statistically supported within-animal association after full adjustment.

### Build the integrated pair-evidence summary

The DE results, target-database support, and association results are stored in
separate files. Script 10 brings those existing results together so that each
of the 56 measured miRNA–mRNA pairs can be reviewed in one row.

The script does not run another statistical test, calculate a combined score,
or claim that one database source is equivalent to another. It preserves the
miRTarBase and TargetScanMouse evidence separately and applies transparent
description rules:

1. A fully supported regulatory pattern would require inverse miRNA and mRNA
   DE directions, a negative fully adjusted association, and association
   FDR < 0.10.
2. A directionally compatible but unconfirmed pair has inverse DE directions
   and a negative adjusted association slope, but its association does not
   pass FDR < 0.10.
3. An inverse-DE-only pair changes in opposite directions, but its fully
   adjusted association is not negative.
4. The remaining pairs change in the same DE direction and do not have
   adjusted association FDR support.

These categories organize the evidence; they are not proof of direct
miRNA-mediated regulation.

Run the integration-summary script from the repository root:

```bash
Rscript scripts/multiomics/10_build_integrated_pair_evidence_summary.R
```

> **Script 10 outputs**
>
> - `results/multiomics/integrated_pair_evidence/mirna_mrna_integrated_pair_evidence_summary.csv`
>   contains one row for each of the 56 pairs. It includes feature and gene
>   identifiers, miRNA and mRNA DE statistics, separate target-resource
>   evidence, the primary 43-animal association, the 27-animal MCAO-focused
>   result, the covariate-model comparisons, and the descriptive evidence
>   category.
>
> - `results/multiomics/integrated_pair_evidence/mirna_mrna_directionally_compatible_unconfirmed_pairs.csv`
>   contains only pairs with inverse DE directions and a negative fully
>   adjusted association slope whose association does not pass FDR < 0.10.
>   This is a review list, not a list of statistically confirmed regulatory
>   relationships.
>
> - `results/multiomics/integrated_pair_evidence/mirna_mrna_integrated_pair_evidence_summary.txt`
>   records pair, miRNA, gene, direction, association, target-source, and
>   evidence-category counts and states the interpretation boundary.

Inspect and validate the integrated evidence:

```{r}
#| label: inspect-integrated-pair-evidence
#| eval: false

integrated_evidence_dir <- file.path(
  project_root,
  "results/multiomics/integrated_pair_evidence"
)
integrated_pairs <- read.csv(
  file.path(
    integrated_evidence_dir,
    "mirna_mrna_integrated_pair_evidence_summary.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
compatible_unconfirmed_pairs <- read.csv(
  file.path(
    integrated_evidence_dir,
    "mirna_mrna_directionally_compatible_unconfirmed_pairs.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

c(
  selected_pairs = nrow(integrated_pairs),
  unique_mirnas = length(unique(integrated_pairs$mirna_id)),
  unique_genes = length(unique(integrated_pairs$gene_entrez_id)),
  inverse_de_pairs = sum(integrated_pairs$inverse_logFC_direction),
  compatible_unconfirmed_pairs = nrow(compatible_unconfirmed_pairs),
  association_fdr_supported_pairs = sum(
    integrated_pairs$full_association_fdr_supported
  )
)
table(integrated_pairs$target_evidence_category)
table(integrated_pairs$integrated_evidence_category)

integrated_pair_key <- paste(
  integrated_pairs$mirna_probeset_id,
  integrated_pairs$transcript_cluster_id,
  integrated_pairs$contrast,
  sep = "||"
)

stopifnot(
  nrow(integrated_pairs) == 56L,
  nrow(compatible_unconfirmed_pairs) == 10L,
  !anyDuplicated(integrated_pair_key),
  sum(integrated_pairs$inverse_logFC_direction) == 13L,
  all(compatible_unconfirmed_pairs$inverse_logFC_direction),
  all(compatible_unconfirmed_pairs$negative_adjusted_association),
  !any(compatible_unconfirmed_pairs$adjusted_fdr_lt_0_10),
  !any(integrated_pairs$fully_supported_regulatory_pattern)
)
```

**Expected outcome after running Script 10 on the 56 selected pairs:** 13
pairs have inverse DE directions. Ten of those also have a negative fully
adjusted association slope and are labeled directionally compatible but
unconfirmed; six have miRTarBase-only support and four have
TargetScanMouse-only support. The other three inverse-DE pairs do not have a
negative fully adjusted association. The remaining 43 pairs change in the
same DE direction. No pair has association FDR < 0.10 in either the primary
43-animal model or the 27-animal MCAO-focused model, so none is classified as
a statistically supported regulatory pattern.

### Prepare enrichment inputs

GSEA and over-representation analysis answer different questions. GSEA uses
the complete ranked mRNA result and asks whether genes from a pathway tend to
occur toward the increased or decreased end of that ranking. It does not
require a DE cutoff. Over-representation analysis starts with a selected gene
list and asks whether particular pathways occur more often than expected
relative to a defined background.

Script 11 prepares the inputs without running an enrichment test. The planned
analysis hierarchy is:

1. **Primary pathway analysis — full-ranked GSEA.** Conventional mRNA GSEA
   will use all measured genes ranked by the limma moderated t-statistic for
   each treatment–sex–age contrast. This is the primary method because it
   uses the complete result and does not depend on an arbitrary DE cutoff.
2. **Secondary pathway analysis — formal ORA of 53 genes.** The only formal
   multi-omics ORA list contains the 53 treatment-associated,
   miRNA-target-linked genes. Its background contains the 1,106 measured
   genes that are targets of at least one of the six selected miRNAs. This
   background reflects the genes that could have entered the 53-gene list.
3. **Multi-omics target-set analysis.** Six measured miRNA target sets are
   prepared to test whether the targets of each selected miRNA tend to occur
   toward one end of a complete mRNA GSEA ranking.
4. **Descriptive review only — the 12- and 9-gene lists.** The 12 unique
   inverse-DE genes and 9 unique directionally compatible but unconfirmed
   genes are retained for annotation and transparent review, but they will
   not be used for formal standalone ORA.

The 12- and 9-gene lists are not mathematically invalid, but they are too
small for reliable standalone pathway testing. A result could be driven by
only one or two genes, statistical power would be low, and the identified
pathways could change substantially if a single gene were added or removed.
They are also nested subsets selected from the same 53-gene result. Treating
them as separate formal enrichment analyses would therefore invite
over-interpretation. Keeping them as descriptive lists preserves the evidence
without presenting unstable pathway findings.

The mRNA DE files contain 20,222 uniquely mapped transcript clusters but
20,082 unique Entrez genes. GSEA requires one ranking value per gene. For the
85 Entrez genes represented by more than one transcript cluster, Script 11
selects the feature with the highest average expression and uses the
transcript-cluster ID to resolve an exact tie. This selection is made once
without using a contrast-specific t-statistic or p-value, so it cannot favor a
different transcript cluster in each contrast. This rule and its counts are
documented here rather than written to a separate transcript-cluster audit
file. The resulting `01_all_measured_mrna_genes.csv` retains the selected
transcript-cluster ID, average expression, number of features mapped to each
gene, and selection rule, so every chosen representative remains traceable.

Run the enrichment-input script from the repository root:

```bash
Rscript scripts/multiomics/11_prepare_enrichment_inputs.R
```

> **Script 11 outputs**
>
> - `results/multiomics/enrichment_inputs/gsea/` contains an annotated CSV
>   and a two-column Entrez-ID `.rnk` file for each of the three contrasts.
>   Every file contains 20,082 unique genes ranked from the largest to the
>   smallest moderated t-statistic.
>
> - `results/multiomics/enrichment_inputs/ora/` contains only
>   `formal_ora_selected_53_target_linked_genes.csv`. This is the sole formal
>   multi-omics ORA list. Repeated miRNA links to the same gene are summarized
>   rather than counted as separate genes.
>
> - `results/multiomics/enrichment_inputs/descriptive_review_lists/` contains
>   the 12-gene inverse-DE list, the 9-gene directionally compatible but
>   unconfirmed list, and a README explaining why these small lists are for
>   descriptive annotation only and are not formal ORA inputs.
>
> - `results/multiomics/enrichment_inputs/backgrounds/` contains the
>   following progressively narrower, numbered backgrounds:
>
>   1. `01_all_measured_mrna_genes.csv`: all 20,082 unique measured mRNA
>      genes, with one selected transcript cluster per Entrez gene.
>   2. `02_all_measured_mirna_target_genes.csv`: the 12,742 measured genes
>      linked to at least one measured miRNA through miRTarBase and/or
>      TargetScanMouse.
>   3. `03_candidate_mirna_target_gene_background.csv`: the 1,106 measured
>      genes targeted by at least one of the six miRNAs selected by Script 10.
>      This is the statistical background for formal ORA of the 53-gene list.
>
> - `results/multiomics/enrichment_inputs/mirna_target_gene_sets/` contains
>   1,159 unique measured miRNA–gene links, a target-set size summary, and an
>   Entrez-ID `.gmt` file containing one gene set for each of the six selected
>   miRNAs.
>
> - `results/multiomics/enrichment_inputs/enrichment_input_summary.txt`
>   records all feature-selection, ranking, background, selected-list, and
>   target-set counts and states that no enrichment test has yet been run.

Inspect and validate the enrichment inputs:

```{r}
#| label: inspect-enrichment-inputs
#| eval: false

enrichment_input_dir <- file.path(
  project_root,
  "results/multiomics/enrichment_inputs"
)
gsea_files <- list.files(
  file.path(enrichment_input_dir, "gsea"),
  pattern = "[.]csv$",
  full.names = TRUE
)
gsea_rankings <- lapply(
  gsea_files,
  read.csv,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

all_measured_background <- read.csv(
  file.path(
    enrichment_input_dir,
    "backgrounds/01_all_measured_mrna_genes.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
all_target_linked_background <- read.csv(
  file.path(
    enrichment_input_dir,
    "backgrounds/02_all_measured_mirna_target_genes.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
candidate_target_background <- read.csv(
  file.path(
    enrichment_input_dir,
    "backgrounds/03_candidate_mirna_target_gene_background.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
selected_ora_genes <- read.csv(
  file.path(
    enrichment_input_dir,
    "ora/formal_ora_selected_53_target_linked_genes.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
inverse_review_genes <- read.csv(
  file.path(
    enrichment_input_dir,
    paste0(
      "descriptive_review_lists/",
      "inverse_de_12_unique_genes_descriptive.csv"
    )
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
compatible_review_genes <- read.csv(
  file.path(
    enrichment_input_dir,
    paste0(
      "descriptive_review_lists/",
      "directionally_compatible_unconfirmed_",
      "9_unique_genes_descriptive.csv"
    )
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
target_set_sizes <- read.csv(
  file.path(
    enrichment_input_dir,
    "mirna_target_gene_sets/candidate_mirna_measured_target_gene_set_sizes.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

stopifnot(
  length(gsea_rankings) == 3L,
  all(vapply(gsea_rankings, nrow, integer(1)) == 20082L),
  all(vapply(
    gsea_rankings,
    function(data) !anyDuplicated(data$ENTREZID),
    logical(1)
  )),
  nrow(all_measured_background) == 20082L,
  nrow(all_target_linked_background) == 12742L,
  nrow(candidate_target_background) == 1106L,
  all(
    all_target_linked_background$gene_entrez_id %in%
      all_measured_background$gene_entrez_id
  ),
  all(
    candidate_target_background$gene_entrez_id %in%
      all_target_linked_background$gene_entrez_id
  ),
  nrow(selected_ora_genes) == 53L,
  nrow(inverse_review_genes) == 12L,
  nrow(compatible_review_genes) == 9L,
  all(
    selected_ora_genes$gene_entrez_id %in%
      candidate_target_background$gene_entrez_id
  ),
  all(
    inverse_review_genes$gene_entrez_id %in%
      selected_ora_genes$gene_entrez_id
  ),
  all(
    compatible_review_genes$gene_entrez_id %in%
      inverse_review_genes$gene_entrez_id
  ),
  nrow(target_set_sizes) == 6L
)
```

**Expected outcome after running Script 11:** each of the three mRNA
contrasts has a complete 20,082-gene GSEA ranking. The gene-level selection
selection rule documented above resolves 85 genes represented by multiple
transcript clusters and excludes 140 lower-expression or later tied
transcript-cluster rows. The three numbered backgrounds contain 20,082
measured genes, 12,742 measured miRNA-target genes, and 1,106 measured targets
of the six selected miRNAs, respectively. The sole formal ORA list contains
53 unique genes. The separate descriptive review lists contain 12 and 9
unique genes and are explicitly excluded from formal standalone ORA. The six
miRNA target sets contain 24, 253, 139, 29, 557, and 157 measured genes,
respectively. These files contain no enrichment p-values or pathway
conclusions; they are validated inputs for the next testing step.

### Run mRNA GSEA

Script 12 performs GSEA only. It does not use the 53-gene ORA list or its
1,106-gene background, and it does not run an ORA test. For each of the three
contrasts, it uses the complete list of 20,082 measured mRNA genes ranked by
the moderated t-statistic from the `treatment + sex + age` limma model. This
complete ranked list is also the tested gene universe; GSEA does not require
a separate background file.

Two official [Mouse MSigDB
2025.1](https://www.gsea-msigdb.org/gsea/msigdb/mouse/collections.jsp)
collections are tested:

- **Mouse Hallmark (MH)** contains 50 broad gene sets representing
  well-defined biological states and processes. It provides a compact
  overview with less pathway redundancy.
- **Mouse canonical pathways (M2:CP)** provides more detailed pathways from
  curated pathway databases. It gives broader coverage but contains many
  biologically overlapping gene sets.

The Entrez-ID GMT files are stored locally under
`resources/pathway_gene_sets/msigdb_2025.1.Mm/`. Their release, source URLs,
download date, and SHA-256 checksums are recorded in `SOURCE_METADATA.md`.
Script 12 checks the file checksums before testing, which prevents an
unnoticed pathway-file change from altering the results.

GSEA is run with `fgsea::fgseaMultilevel`. A pathway is tested only when
15–500 of its genes occur in the measured ranking. Hallmark and canonical
pathways are analyzed separately for each contrast, so FDR correction is
performed within each of the six contrast-by-collection analyses. The
significance threshold is FDR < 0.05.

Each ranking contains five exact t-statistic tie groups involving ten genes.
Script 12 preserves the deterministic rank order created by Script 11 and
applies a minute numerical decrement only within each tie. The largest
adjustment is approximately \(3.1 \times 10^{-12}\), far smaller than a
biologically meaningful t-statistic difference. This prevents arbitrary
ordering by the GSEA software without changing the substantive ranking.

The normalized enrichment score (NES) describes the direction:

- A **positive NES** means that pathway genes tend to occur near the positive
  end of the t-statistic ranking—that is, higher in the first group named in
  the contrast.
- A **negative NES** means that pathway genes tend to occur near the negative
  end—that is, lower in the first group and relatively higher in the second
  group named in the contrast.

For example, a negative NES in `MCAO3hr_vs_MCAO1hr` means that the pathway
tends toward lower expression at MCAO3hr than at MCAO1hr. It does not mean
that every gene in the pathway decreased.

Run GSEA from the repository root:

```bash
Rscript scripts/multiomics/12_run_mrna_gsea.R
```

> **Script 12 outputs**
>
> - `contrast_results/` contains one complete GSEA result CSV for each
>   contrast. Each row reports the pathway, measured pathway size,
>   enrichment score, NES, p-value, FDR, direction, and leading-edge genes.
>
> - `combined_gsea_results.csv` combines all tested pathways and contrasts
>   into one file.
>
> - `significant_gsea_results_fdr_0.05.csv` contains the rows meeting
>   collection-specific FDR < 0.05.
>
> - `leading_edge_genes.csv` expands the leading-edge genes from significant
>   pathways to one pathway–gene row at a time. A gene can occur repeatedly
>   because pathways overlap.
>
> - `pathway_collection_summary.csv` records how many pathways were tested
>   and significant in each contrast and collection.
>
> - `figures/gsea_pathway_summary.png` shows up to ten pathways with the
>   smallest FDR in each contrast and collection. Red indicates positive NES;
>   blue indicates negative NES.
>
> - `gsea_run_summary.txt` records the method, parameters, pathway release,
>   checksums, tie handling, package version, and result counts.

Inspect and validate the GSEA results:

```{r}
#| label: inspect-mrna-gsea-results
#| eval: false
#| fig-cap: "Top FDR-significant mRNA GSEA pathways by contrast and collection."
#| fig-width: 16
#| fig-height: 12

gsea_result_dir <- file.path(
  project_root,
  "results/multiomics/enrichment_results/gsea"
)
gsea_results <- read.csv(
  file.path(gsea_result_dir, "combined_gsea_results.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
significant_gsea_results <- read.csv(
  file.path(
    gsea_result_dir,
    "significant_gsea_results_fdr_0.05.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
gsea_collection_summary <- read.csv(
  file.path(gsea_result_dir, "pathway_collection_summary.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

stopifnot(
  nrow(gsea_results) == 3273L,
  nrow(significant_gsea_results) == 1474L,
  nrow(gsea_collection_summary) == 6L,
  all(is.finite(gsea_results$normalized_enrichment_score)),
  all(is.finite(gsea_results$p_value)),
  all(is.finite(gsea_results$FDR)),
  all(gsea_results$FDR >= 0 & gsea_results$FDR <= 1),
  all(significant_gsea_results$FDR < 0.05)
)

knitr::include_graphics(
  file.path(
    gsea_result_dir,
    "figures/gsea_pathway_summary.png"
  )
)
```

> **Observed GSEA result counts**
>
> - `MCAO1hr_vs_Sham`: 39 of 50 Hallmark sets and 497 of 1,041 tested
>   canonical pathways met FDR < 0.05.
> - `MCAO3hr_vs_Sham`: 15 Hallmark sets and 259 canonical pathways met
>   FDR < 0.05.
> - `MCAO3hr_vs_MCAO1hr`: 31 Hallmark sets and 633 canonical pathways met
>   FDR < 0.05.
> - Across the six analyses, 1,474 of 3,273 pathway-test rows met FDR < 0.05.

The 1,474 significant rows are not 1,474 independent biological discoveries.
Many canonical pathways share genes or describe different levels of the same
process, and the same pathway can appear in more than one contrast. Script 13
therefore organizes dependent pathways, inspects their leading-edge genes, and
compares their directions across contrasts. These are mRNA pathway results
only; they do not by themselves demonstrate miRNA regulation, causality, or
mediation.

### Summarize and reduce GSEA redundancy

Script 13 summarizes the significant results from Script 12 without rerunning
GSEA and without performing ORA. It uses `fgsea::collapsePathways()` separately
within each contrast and MSigDB collection. This method asks whether a
significant pathway provides enrichment information that is sufficiently
independent of a stronger pathway. If it does not, the pathway is retained in
the audit map but assigned to the stronger representative pathway. Therefore,
no significant pathway is silently discarded.

This is a statistical redundancy reduction rather than a manual grouping
based only on similar pathway names. The dependency p-value threshold is 0.05,
200 permutations are used, and a fixed random seed makes the mapping
reproducible. The original GSEA p-values, FDR values, and NES values are not
recalculated or replaced. Hallmark and canonical pathways also remain
separate because Script 12 treated them as separate FDR families.

A **representative pathway** is a concise label for a group of statistically
dependent results. It should not be read as proof that only the process named
by that representative is active. The accompanying redundancy map must be
used to see every significant pathway assigned to it.

Script 13 also summarizes the **leading-edge genes**. These are the genes that
contribute most strongly to a pathway's enrichment signal. A gene that appears
in the leading edges of several representative pathways may help explain a
shared expression pattern, but repeated occurrence does not prove that the
gene regulates those pathways.

Run the GSEA-summary script from the repository root:

```bash
Rscript scripts/multiomics/13_summarize_mrna_gsea_results.R
```

> **Script 13 outputs**
>
> - `representative_pathways.csv` contains one row for each retained
>   representative in each contrast and collection, together with its NES,
>   FDR, direction, group size, and assigned pathway members.
>
> - `pathway_redundancy_map.csv` preserves all 1,474 significant pathway rows
>   and records whether each pathway is a representative or which
>   representative it was assigned to.
>
> - `representative_leading_edge_genes.csv` expands the leading edges of the
>   representative pathways to one pathway–gene row at a time.
>
> - `leading_edge_driver_gene_summary.csv` counts how many representative
>   pathway leading edges contain each gene. These counts are prioritization
>   aids, not new gene-level significance tests.
>
> - `cross_contrast_representative_pathways.csv` places NES, FDR,
>   significance, and representative status for all three contrasts side by
>   side. It identifies pathways with the same direction or a direction
>   reversal across significant contrasts.
>
> - `redundancy_reduction_summary.csv` records the number and direction of
>   significant and representative pathways for each contrast and
>   collection.
>
> - `figures/representative_pathway_summary.png` shows up to eight
>   representatives with the smallest FDR in each contrast and collection.
>
> - `figures/cross_contrast_pathway_heatmap.png` compares NES values for 36
>   shared representatives. Faded cells are not significant at FDR < 0.05,
>   and asterisks identify contrasts in which the pathway was retained as a
>   representative.
>
> - `gsea_redundancy_summary.txt` records the method, parameters, reduction
>   counts, output locations, and interpretation boundaries.

Inspect and validate the redundancy-reduced results:

```{r}
#| label: inspect-gsea-redundancy-results
#| eval: false

gsea_reduced_dir <- file.path(
  project_root,
  paste0(
    "results/multiomics/enrichment_results/gsea/",
    "redundancy_reduced"
  )
)
representative_pathways <- read.csv(
  file.path(gsea_reduced_dir, "representative_pathways.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
pathway_redundancy_map <- read.csv(
  file.path(gsea_reduced_dir, "pathway_redundancy_map.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
representative_leading_edges <- read.csv(
  file.path(
    gsea_reduced_dir,
    "representative_leading_edge_genes.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
leading_edge_driver_genes <- read.csv(
  file.path(
    gsea_reduced_dir,
    "leading_edge_driver_gene_summary.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
cross_contrast_pathways <- read.csv(
  file.path(
    gsea_reduced_dir,
    "cross_contrast_representative_pathways.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
redundancy_reduction_summary <- read.csv(
  file.path(
    gsea_reduced_dir,
    "redundancy_reduction_summary.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

stopifnot(
  nrow(pathway_redundancy_map) == 1474L,
  nrow(representative_pathways) == 579L,
  nrow(representative_leading_edges) == 29796L,
  nrow(leading_edge_driver_genes) == 13055L,
  nrow(cross_contrast_pathways) == 342L,
  nrow(redundancy_reduction_summary) == 6L,
  sum(pathway_redundancy_map$is_representative) == 579L,
  all(
    paste(
      pathway_redundancy_map$contrast,
      pathway_redundancy_map$collection_id,
      pathway_redundancy_map$representative_pathway
    ) %in%
      paste(
        representative_pathways$contrast,
        representative_pathways$collection_id,
        representative_pathways$pathway
      )
  ),
  all(representative_pathways$FDR < 0.05),
  sum(
    cross_contrast_pathways$cross_contrast_pattern ==
      "direction_reversal"
  ) == 239L
)
```

Display the redundancy-reduced representative-pathway figure:

```{r}
#| label: display-representative-gsea-pathways
#| eval: false
#| fig-cap: "Redundancy-reduced representative mRNA GSEA pathways."
#| fig-width: 16
#| fig-height: 12

knitr::include_graphics(
  file.path(
    gsea_reduced_dir,
    "figures/representative_pathway_summary.png"
  )
)
```

Display the cross-contrast direction heatmap:

```{r}
#| label: display-cross-contrast-gsea-heatmap
#| eval: false
#| fig-cap: "Cross-contrast NES patterns for shared representative pathways."
#| fig-width: 11
#| fig-height: 14

knitr::include_graphics(
  file.path(
    gsea_reduced_dir,
    "figures/cross_contrast_pathway_heatmap.png"
  )
)
```

> **Observed redundancy-reduction results**
>
> - The 1,474 significant pathway rows were reduced to 579
>   contrast-specific representative rows.
> - Canonical pathways were reduced from 497 to 182 representatives in
>   `MCAO1hr_vs_Sham`, from 259 to 102 in `MCAO3hr_vs_Sham`, and from 633 to
>   213 in `MCAO3hr_vs_MCAO1hr`. This corresponds to approximately 61–66%
>   reduction.
> - Hallmark results changed much less: 39 to 38, 15 to 13, and 31 to 31
>   representatives, respectively. This is expected because Hallmark was
>   designed as a compact, less-redundant collection.
> - The representatives comprise 342 unique pathways across contrasts. Of
>   these, 37 were significant in one contrast, 191 in two contrasts, and 114
>   in all three.
> - A total of 239 representatives showed a direction reversal: positive NES
>   in at least one significant contrast and negative NES in another.

The cross-contrast results support a time-dependent pattern. Many immune,
stress-response, growth-signaling, RNA-processing, and cell-maintenance
programs were enriched toward higher expression in `MCAO1hr_vs_Sham` but
toward lower expression in `MCAO3hr_vs_MCAO1hr`. RNA processing, translation,
cell-cycle, DNA-repair, and oxidative-phosphorylation themes were also
predominantly lower in `MCAO3hr_vs_Sham`. This pattern is consistent with a
broad early response followed by lower expression of many of the same
programs after reperfusion and recovery. It remains a pathway-level
association and does not establish a causal time course or miRNA-mediated
regulation.

### Run formal multi-omics ORA

Script 14 performs the formal ORA separately from GSEA. It asks whether the 53
selected genes occur in particular pathways more often than expected among
the genes that could have entered the selected list.

The 53 genes come from the 56 unique miRNA–mRNA pairs identified in
`MCAO3hr_vs_MCAO1hr`. Every pair met miRNA FDR < 0.10 and mRNA FDR < 0.10 and
was supported as a measured target relationship by miRTarBase and/or
TargetScanMouse. Three genes occurred in two pairs, so the 56 pairs correspond
to 53 unique genes.

The statistical background contains 1,106 unique measured genes targeted by
at least one of the same six selected miRNAs. This is more appropriate than
using all 20,082 measured mRNA genes because genes without a target link to
one of these six miRNAs could not have entered the 53-gene list under this
selection design.

For each pathway, Script 14 constructs four counts: selected genes inside the
pathway, selected genes outside the pathway, non-selected background genes
inside the pathway, and non-selected background genes outside the pathway. A
one-sided Fisher exact test then asks whether the proportion inside the
pathway is greater for the 53 selected genes than for the complete 1,106-gene
background.

Mouse Hallmark and canonical pathways are tested separately. A pathway is
eligible when 5–500 of its genes occur in the 1,106-gene background.
Benjamini–Hochberg FDR correction is applied separately within the Hallmark
and canonical collections, and FDR < 0.05 is the formal significance
threshold. Raw p < 0.05 is retained for transparent review but is not treated
as formal evidence without FDR support.

Run ORA from the repository root:

```bash
Rscript scripts/multiomics/14_run_multiomics_ora.R
```

> **Script 14 outputs**
>
> - `collection_results/mouse_hallmark_ora_results.csv` contains all eligible
>   Hallmark tests.
>
> - `collection_results/mouse_canonical_pathways_ora_results.csv` contains
>   all eligible canonical-pathway tests.
>
> - `combined_ora_results.csv` combines both collections. Each row includes
>   the selected and background pathway counts, fold enrichment, odds ratio,
>   raw p-value, FDR, and overlapping selected genes.
>
> - `significant_ora_results_fdr_0.05.csv` contains only formally significant
>   pathways. It currently contains a header but no result rows because no
>   pathway met FDR < 0.05.
>
> - `nominal_p_lt_0.05_ora_results.csv` retains the two raw-p-value findings
>   for exploratory review. They must not be described as statistically
>   significant ORA results.
>
> - `pathway_selected_gene_overlaps.csv` expands each selected pathway–gene
>   overlap and retains the gene's selected miRNAs and mRNA DE statistics.
>
> - `pathway_collection_summary.csv` records the source, eligible, nominal,
>   and FDR-significant pathway counts for each collection.
>
> - `figures/ora_top_pathways.png` displays the ten pathways with the smallest
>   raw p-values per collection. Its subtitle explicitly states that none met
>   FDR < 0.05.
>
> - `ora_run_summary.txt` records the selected list, background, method,
>   pathway-size rule, MSigDB version, checksums, thresholds, and result
>   counts.

Inspect and validate the formal ORA results:

```{r}
#| label: inspect-formal-multiomics-ora
#| eval: false

ora_result_dir <- file.path(
  project_root,
  "results/multiomics/enrichment_results/ora"
)
ora_results <- read.csv(
  file.path(ora_result_dir, "combined_ora_results.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
significant_ora_results <- read.csv(
  file.path(
    ora_result_dir,
    "significant_ora_results_fdr_0.05.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
nominal_ora_results <- read.csv(
  file.path(
    ora_result_dir,
    "nominal_p_lt_0.05_ora_results.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
ora_gene_overlaps <- read.csv(
  file.path(
    ora_result_dir,
    "pathway_selected_gene_overlaps.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
ora_collection_summary <- read.csv(
  file.path(ora_result_dir, "pathway_collection_summary.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

stopifnot(
  nrow(ora_results) == 393L,
  nrow(significant_ora_results) == 0L,
  nrow(nominal_ora_results) == 2L,
  nrow(ora_gene_overlaps) == 222L,
  nrow(ora_collection_summary) == 2L,
  all(is.finite(ora_results$p_value)),
  all(is.finite(ora_results$FDR)),
  all(ora_results$p_value >= 0 & ora_results$p_value <= 1),
  all(ora_results$FDR >= 0 & ora_results$FDR <= 1),
  all(nominal_ora_results$p_value < 0.05),
  all(nominal_ora_results$FDR >= 0.05)
)
```

Display the ORA summary figure:

```{r}
#| label: display-formal-multiomics-ora
#| eval: false
#| fig-cap: "Top-ranked formal ORA pathways; no pathway met FDR < 0.05."
#| fig-width: 14
#| fig-height: 8

knitr::include_graphics(
  file.path(
    ora_result_dir,
    "figures/ora_top_pathways.png"
  )
)
```

> **Observed formal ORA results**
>
> - Forty-four Hallmark pathways and 349 canonical pathways were eligible,
>   giving 393 formal tests.
> - `HALLMARK_SPERMATOGENESIS` contained 2 of the 53 selected genes among 7
>   background genes in that set. Its fold enrichment was 5.96, raw p-value
>   was 0.0405, and FDR was 1.00. The overlapping genes were `Hspa4l` and
>   `Zc3h14`.
> - `WP_MRNA_PROCESSING` contained 5 selected genes among 40 background genes
>   in that set. Its fold enrichment was 2.61, raw p-value was 0.0382, and FDR
>   was 1.00. The overlapping genes were `Atxn1`, `Rsrc2`, `Srp14`, `Srpk2`,
>   and `Syncrip`.
> - No Hallmark or canonical pathway met FDR < 0.05.

The formal conclusion is that the 53-gene list did not provide statistically
significant evidence of pathway over-representation relative to its
1,106-gene selection background. The two raw-p-value findings are exploratory
only and should not be presented as confirmed pathway enrichment. This
negative ORA result does not contradict the GSEA results: GSEA used the full
20,082-gene mRNA rankings and could detect coordinated shifts spread across
many genes, whereas ORA tested whether a much smaller, preselected 53-gene
list contained an unusually large number of pathway members.

<hr class="section-divider">

# Record software versions

The final chunk records the R version, operating system, and loaded package
versions used in the rendered workflow. This information helps reproduce the
analysis and diagnose future package-version differences.

```{r}
#| label: session-information

sessionInfo()
```

**Expected outcome:** a compact record of the active R session and package
versions. This step reads the environment only and does not alter data or
results.
